---
title: "The Anscombe Exercise"
author: "Jilly MacKay"
date: "21 November 2017"
output:
  html_document:
    keep_md: true
    theme: spacelab
    highlight: haddock
    toc: true 
    toc_float: true
    toc_depth: 3
---
##About This Exercise
**Exercise Type:** Confidence Builder

*Confidence Builders are designed to get you more comfortable with R. We'll go through bits of code more or less step-by-step, and describe the reasoning behind certain actions. Remember - it's not cheating to copy code from this document and paste it into your R Studio console!*

**Exercise Suitability:** Beginner

*If you have R Studio installed on your machine and you know what an object is, you can work through this material.*

#The Anscombe Exercise
This exercise will take you through a really fun little dataset called the *Anscombe Quartet*. I've been trying to teach this exercise using data I've made up myself, and it's never quite worked. Imagine my delight when I heard that Francis Anscombe had already created the perfect dataset for me! At least, according to Wikipedia, no one knows exactly how he came up with it.

The **aim** of this exercise is to make you more comfortable with entering code in R. 

##Learning Outcomes
By the end of this workshop you should be able to . . .

* Provide descriptive statistics for data
* Run a simple regression on data
* Change arguments in a function
* Create visualisations in R


#The R Environment
Here are the packages you'll need for this exercise. (Remember, if you don't have one of these packages you can install it with the command: `install.packages(packagename)`
```{r}
library(ggplot2)
library(gridExtra)
library(pastecs)
library(tidyr)
library(dplyr)
```


#The Anscombe Dataset
The Anscombe dataset is really interesting and comes pre-loaded as part of the package `datasets`. This is a package which comes as standard in any R installation and so we can look at it immediately just by typing the name of the dataset into R Studio. 

R Studio will show us all rows of this dataset because it's quite small (only 11 rows). If we had a larger dataset, it would cut off after about 10 rows.

```{r}
anscombe 
```



#Running Descriptive Statistics
Let's start by exploring the means and medians of all these variables. There is a very handy command `stat.desc` from the package`pastecs` which can do this for us. 

(Note - there are LOADS of packages which give you descriptive stats. Check out [this statmethods page](https://www.statmethods.net/stats/descriptives.html) for more examples. I just personally like `pastecs`.)

```{r}
stat.desc(anscombe)
```

For each variable in the array, `stat.desc` gives us:

* `nbr.val` Number of values within the variable
* `nbr.null` Number of null values
* `nbr.na` Number of missing values
* `min` Minimum value
* `max` Maximum value
* `range` Maximum-minimum value
* `sum` Sum of all non-missing values
* `median` The median value
* `mean` The mean value
* `SE.mean` The standard error of the mean
* `CI.mean.0.95` The 95% confidence intervals of this mean
* `var` The variance of this variable
* `std.dev` The standard deviation of the mean
* `coef.var` The variation coefficient (standard deviation/mean)

With these descriptive statistics, we might want to start writing our interpretation of the data. For example ...

> The average score of *x* was 9 ±3.3 standard deviations.

###Changing Arguments in a Function

You might decide you don't want to bother with *all* the elements of `stat.desc` and so you could explore how to modify the command by asking R `help("stat.desc")` which will show you the help documentation for the function. Note that it talks about *arguments*. 

```{r}
stat.desc(anscombe, basic=FALSE, norm=FALSE, p=0.99)
```

By setting the arguments `basic=FALSE` and `norm=FALSE`, we've told R we only care about the information that the `desc` argument gives us. Note, because we only want the `desc` argument we don't need to bother typing it in. It's set to `true` by default. You'd get the same result by typing `stat.desc(anscombe, basic=FALSE, desc=TRUE, norm=FALSE, p=0.99)` which you can try yourself. We've also changes the significance level of the confidence intervals to a p level of 0.99. 

In this example, there's no real harm in getting all the information `stat.desc` can give us, but when we start plotting the data we will want to play about with the arguments in the `ggplot` command, so that's why you're seeing them here!

#Run a Regression
We're going to run four regressions for this dataset (all four *x*s against all 4 *y*s) and explore them. This will be done using R's inbuilt stats package, so there's no need to load any specific package into our library. 

```{r}
regression1 <- lm (y1~x1, data=anscombe)
regression2 <- lm (y2~x2, data=anscombe)
regression3 <- lm (y3~x3, data=anscombe)
regression4 <- lm (y4~x4, data=anscombe)
summary (regression1)
summary (regression2)
summary (regression3)
summary (regression4)
```

Here we might want to report these findings. 

> There was a significant, positive relationship between *x* and *y* (F^1,9^=18, p=0.002) in which *x* explained approximately 63% of the variation observed in *y*.

Now you're probably thinking "Great! Let's write up the paper and go home!"

#Visualising Data
Everything we've seen so far has shown us that we have identical relationships between *x* and *y* in the Anscombe dataset. So here's what we should have done first . . . 

We're going to use `ggplot` for this. (We're also going to lean very heavily on [Ian and Stella's great ggplot tutorial](https://ianhandel.github.io/plotting-with-r/index.html) which I recommend if you want to learn more about `ggplot`). We'll walk through building the first chart and then build the others in one go. 

##Understanding ggplot layers

```{r}
ggplot (data=anscombe, aes(x=x1, y=y1))
```

The `ggplot` command first needs to know what *data* to use, (the `data` argument) and then what *aesthetics* ( the `aes` argument). But you will have spotted pretty quickly that there's no data in there yet! We need to start layering information into the `ggplot` command. The two most important types of layers are the geometric elements (`geoms`) and statistical transformations (`stats`). 

We'll begin by adding a layer of points with the `geom_point` function.

```{r}
ggplot (data=anscombe, aes(x=x1, y=y1)) +
  geom_point()
```

Now we're getting somewhere! Let's add a statistical layer. 

```{r}
ggplot (data=anscombe, aes(x=x1, y=y1)) +
  geom_point() +
  stat_smooth (method="lm", se=FALSE)
```

This looks like a proper chart! Still, your research methods teacher will probably shout at you if you don't have the x and y axis start at zero . . . 

```{r}
ggplot (data=anscombe, aes(x=x1, y=y1)) +
  geom_point() +
  stat_smooth(method="lm", se=FALSE) +
  expand_limits(x=0, y=0)
```

And she'll also shout at you for not having the x and y axis share sensible scales 

```{r}
ggplot (data=anscombe, aes(x=x1, y=y1)) +
  geom_point() +
  stat_smooth(method="lm", se=FALSE) +
  expand_limits(x=0, y=0) +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  scale_y_continuous(breaks = seq(0, 10, 2))
```

And why is the background grey? You know that's a waste of ink if you print it out. 

```{r}
ggplot(data=anscombe, aes(x=x1, y=y1)) +
  geom_point() +
  stat_smooth(method="lm", se=FALSE) +
  theme_bw() +
  expand_limits(x=0, y=0) +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  scale_y_continuous(breaks = seq(0, 10, 2))
```

Maybe you need a title, and better axis labels too?

```{r}
ggplot(data=anscombe, aes(x=x1, y=y1)) +
  geom_point() +
  stat_smooth(method="lm", se=FALSE) +
  theme_bw() +
  expand_limits(x=0, y=0) +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  scale_y_continuous(breaks = seq(0, 10, 2)) +
  ggtitle("Anscombe Dataset 1") +
  xlab("Variable X1") +
  ylab("Variable Y1")
```

##Building ggplots
Personally, I like to make my ggplots objects like so . . . 
```{r}
Anscombe1 <- ggplot(data=anscombe, aes(x=x1, y=y1)) +
  geom_point() +
  stat_smooth(method="lm", se=FALSE) +
  theme_bw() +
  expand_limits(x=0, y=0) +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  scale_y_continuous(breaks = seq(0, 10, 2)) +
  ggtitle("Anscombe Dataset 1") +
  xlab("Variable X1") +
  ylab("Variable Y1")
```

Then I can call them back much more easily if I want to look at them again

```{r}
Anscombe1
```

And now I can combine these into a grid by taking the first plot `Anscombe1` and adding new column mappings e.g.  `aes(x=x2, y=y2)` and labelling e.g. `ggtitle("Anscombe Dataset 2")`. 
```{r}
Anscombe2 <- Anscombe1 +
  aes(x=x2, y=y2) +
  ggtitle("Anscombe Dataset 2") + xlab("Variable X2") + ylab("Variable Y2")
Anscombe3 <- Anscombe1 +
  aes(x=x3, y=y3) +
  ggtitle("Anscombe Dataset 3") + xlab("Variable X3") + ylab("Variable Y3")
Anscombe4 <- Anscombe1 +
  aes(x=x4, y=y4) +
  ggtitle("Anscombe Dataset 4") + xlab("Variable X4") + ylab("Variable Y4")

#And now let's put them all together!

grid.arrange(Anscombe1, Anscombe2, Anscombe3, Anscombe4, top="This is why you always visualise data first!")
```


Lesson learned, eh?!

#Where Do We Go From Here?
If you spend enough time on it, you can create truly beautiful visualisations in `ggplot2`. Play about with this code and see what you can do with it - share your best attempts! 


I want to create a chart with all the sets overlaid, so I need all the x values in one column, all the y columns in another, and a new factor to tell me which set it belonged to. To change the shape of the data I'm going to use the `tidyr` and `dplyr` packages to do this. 

```{r}
LongAnscombe <- anscombe %>%
  mutate(observation=seq_len(n()))%>%
  gather(key, value, -observation)%>% 
  separate(key, c("variable", "set") , 1 , convert=TRUE)%>%
  mutate(set=factor(set))%>%
  spread(variable, value)
LongAnscombe
```

Now I want to plot it!

(I quite like [this colour cheatsheet](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf) for the `scale_colour_manual` function)

```{r}
AnscombeQuartet <- ggplot (data=LongAnscombe,
                           aes (x=x, y=y, color=set, shape=set)) +
  geom_point() +
  scale_colour_manual(name = "Quartet Set",
                      labels = c("One", "Two", "Three", "Four"),
                      values=c("steelblue4", "springgreen4", "slateblue4", "violetred4")) +
  scale_shape_manual(name = "Quartet Set",
                     labels=c("One", "Two", "Three", "Four"),
                     values=c(16, 17, 18, 15)) +
  stat_smooth(method = "lm", se = FALSE, lwd = 0.5) +
  theme_bw() +
  expand_limits (x=0, y=0) +
  ggtitle ("The Anscombe Quartet") +
  xlab ("The X Variable") +
  ylab ("The Y Variable") +
  facet_grid(. ~ set)
AnscombeQuartet
```

Start messing about with this bit of `ggplot` code to produce your prettiest examples of the Anscombe Quartet. If you feel like you've spent enough time visualising them, here's what I suggest you do next...

`install.packages("datasauRus")`
```{r, echo=FALSE}
library(datasauRus)
```
And then you can do this . . .

```{r}
ggplot(datasaurus_dozen, aes(x=x, y=y, colour=dataset))+
  geom_point()+
  theme_void()+
  theme(legend.position = "none")+
  facet_wrap(~dataset, ncol=3)
```

And check out these:
```{r}
stat.desc(datasaurus_dozen_wide, basic=FALSE, norm=FALSE)
```

Got the message yet? ;)
